---
title: "Economic recovery by educational attainment"
author: "Andrew Ba Tran"
date: "July 12, 2016"
output: html_document
---

This is the methodology used for the Trend CT story: [Economic recovery leaving behind those who did not go to college](http://trendct.org/2016/07/13/countrys-economic-recovery-leaving-behind-those-who-did-not-go-to-college/).

Visit the [repo](https://github.com/trendct-data/job-gap/) for the [data](https://github.com/trendct-data/job-gap/tree/master/data) used in this analysis. (Also, check out the reproducible scripts and data behind many of our other stories in our [central data stories repo](https://github.com/trendct-data))

The data used in this analysis was either copy and pasted and cleaned by hand from the Census or brought in using the Census API.

What’s in this walkthrough

Several charts exploring the national and Connecticut-specific employment figures by educational attainment levels.

```{r setup, message=F, warning=F}
library(dplyr)
library(tidyr)
library(extrafont)
library(ggalt)
library(lubridate)
library(dplyr)
library(tidyr)
library(stringr)
library(scales)
library(gridExtra)
library(grid)
library(ggplot2)
# install.packages("devtools")
#devtools::install_github("hrecht/censusapi")
library("censusapi")

# This file contains my census api key
source("keys.R")

# Sign up for your own key
# http://api.census.gov/data/key_signup.html
```

# National monthly job change

```{r monthly, fig.width=9, fig.height=7, warning=F, message=F}
# Bringing in US data. Curated from http://www.census.gov/cps/data/cpstablecreator.html

us_change <- read.csv("data/us_all_raw.csv")
us_change$Date <- paste0("1-", us_change$Date)
us_change$Date <- dmy(us_change$Date)

colnames(us_change) <- c("Date", "Bachelor's degree or higher", "Associate's degree or some college", "High school or less")

us_long <- mutate(gather(us_change, area, value, -Date),
                      area=factor(area, levels=colnames(us_change)[2:4],
                                  ordered=TRUE))

edu_colors <- c('#66c2a5', '#fc8d62', '#8da0cb')
names(edu_colors) <- colnames(us_change)[2:4]

last_vals <- sapply(colnames(us_change)[2:4], function(x) last(na.exclude(us_change[,x])))
last_date <- tail(us_change$Date)#+1 # doing this ^ wld have made it a double

# jobs change (monthly)

gg <- ggplot(us_long)
gg <- gg + geom_hline(yintercept = 0)
gg <- gg + geom_line(aes(x=Date, y=value, group=area, color=area))
#gg <- gg + geom_rect(aes(xmin=as.numeric(ymd("2007-12-01")), xmax=as.numeric(ymd("2010-01-01")), ymin=-Inf, ymax=Inf))
gg <- gg + annotate("rect", xmin = ymd("2007-12-01"), xmax =ymd("2010-01-01"), ymin = -Inf, ymax = Inf,
                    alpha = .1)

gg <- gg + scale_color_manual(name="", values=edu_colors)
#gg <- gg + scale_y_continuous(label=percent)
gg <- gg + labs(x=NULL, y="Employment change (millions)", title="Job change in recession and recovery in the U.S.",
                subtitle="Workers with more than a high school education gained 11.5 million of the 11.6 million jobs added in the recovery.",
                caption="SOURCE: U.S. Census CPS, Georgetown Center on Education and the Workforce")
gg <- gg + theme_bw(base_family="Helvetica")
gg <- gg + theme(axis.ticks.y=element_blank())
gg <- gg + theme(panel.border=element_blank())
gg <- gg + theme(legend.key=element_blank())
gg <- gg + theme(plot.title=element_text(face="bold", family="Lato Black", size=22))
gg <- gg + theme(plot.caption=element_text(face="bold", family="Lato", size=9, color="gray", margin=margin(t=10, r=80)))
gg <- gg + geom_label(data=us_long, aes(x=ymd("2008-4-01"), y=5, label="Recession", hjust=0),
                      family="Helvetica", lineheight=0.95,
                      size=5, label.size=0, color="#2b2b2b")
gg <- gg + geom_label(data=us_long, aes(x=ymd("2010-4-01"), y=5, label="Recovery", hjust=0),
                      family="Helvetica", lineheight=0.95,
                      size=5, label.size=0, color="#2b2b2b")
gg <- gg + theme(legend.position="none")
gg <- gg + theme(plot.margin = unit(c(1, 15, 1, 1), "lines"))
gg


for (i in 1:length(last_vals)) {
  gg <- gg + annotation_custom(grob=textGrob(paste0(names(last_vals)[i], " (", round(last_vals[i],1), ")"), hjust=-.3,
                                             gp=gpar(fontsize=10, 
                                                     col=edu_colors[names(last_vals)[i]])),
                               xmin=as.numeric(last_date[i]),xmax=as.numeric(last_date[i]),
                               ymin=last_vals[i], ymax=last_vals[i])
}

gb <- ggplot_build(gg)
gt <- ggplot_gtable(gb)

gt$layout$clip[gt$layout$name=="panel"] <- "off"
grid.draw(gt)

```

# Connecticut annual job change


```{r ct_only, fig.width=9, fig.height=7, warning=F, message=F}
# These Census IDs are what we're pulling from the Census API

# B23006_001E - Total
# B23006_002E - Less than high school graduate
# B23006_009E - High school graduate (includes equivalency)
# B23006_016E - Some college or associate's degree
# B23006_023E - Bachelor's degree or higher

# Function from censusapi to pull 2009 data
# Replace `census_key` with your individual API key

data2009 <- getCensus(name="acs5", 
                      vintage=2009,
                      key=census_key, 
                      vars=c("NAME", "B23006_002E",
                             "B23006_009E", "B23006_016E", "B23006_023E"), 
                      region="state:*")

# adding year identifier for future joining

data2009$state <- 2009

# Cleaning up the column names
colnames(data2009) <- c("state", "year", "Less than high school graduate", "High school graduate", "Some college or associate's degree", "Bachelor's degree or higher")

# Some math
data2009$`High school or less` <- data2009$`High school graduate` + data2009$`Less than high school graduate`

# Deleting unnecessary columns
data2009[,3] <- NULL
data2009[,3] <- NULL

# Restructuring the data for joining
data2009 <- data2009 %>%
  gather("category", "total", 3:5)

# Function from censusapi to pull 2010 data
# Most of what flows below is the same as the code above but for a different year
# Probably would be more efficient to write a loop to compile this but it's too late now

data2010 <- getCensus(name="acs5", 
                      vintage=2010,
                      key=census_key, 
                      vars=c("NAME", "B23006_002E",
                             "B23006_009E", "B23006_016E", "B23006_023E"), 
                      region="state:*")
data2010$state <- 2010
colnames(data2010) <- c("state", "year", "Less than high school graduate", "High school graduate", "Some college or associate's degree", "Bachelor's degree or higher")
data2010$`High school or less` <- data2010$`High school graduate` + data2010$`Less than high school graduate`
data2010[,3] <- NULL
data2010[,3] <- NULL

data2010 <- data2010 %>%
  gather("category", "total", 3:5)

data2011 <- getCensus(name="acs5", 
                      vintage=2011,
                      key=census_key, 
                      vars=c("NAME", "B23006_002E",
                             "B23006_009E", "B23006_016E", "B23006_023E"), 
                      region="state:*")
data2011$state <- 2011
colnames(data2011) <- c("state", "year", "Less than high school graduate", "High school graduate", "Some college or associate's degree", "Bachelor's degree or higher")
data2011$`High school or less` <- data2011$`High school graduate` + data2011$`Less than high school graduate`
data2011[,3] <- NULL
data2011[,3] <- NULL

data2011 <- data2011 %>%
  gather("category", "total", 3:5)

data2012 <- getCensus(name="acs5", 
                      vintage=2012,
                      key=census_key, 
                      vars=c("NAME", "B23006_002E",
                             "B23006_009E", "B23006_016E", "B23006_023E"), 
                      region="state:*")
data2012$state <- 2012
colnames(data2012) <- c("state", "year", "Less than high school graduate", "High school graduate", "Some college or associate's degree", "Bachelor's degree or higher")
data2012$`High school or less` <- data2012$`High school graduate` + data2012$`Less than high school graduate`
data2012[,3] <- NULL
data2012[,3] <- NULL

data2012 <- data2012 %>%
  gather("category", "total", 3:5)

data2013 <- getCensus(name="acs5", 
                      vintage=2013,
                      key=census_key, 
                      vars=c("NAME", "B23006_002E",
                             "B23006_009E", "B23006_016E", "B23006_023E"), 
                      region="state:*")

data2013$state <- 2013
colnames(data2013) <- c("state", "year", "Less than high school graduate", "High school graduate", "Some college or associate's degree", "Bachelor's degree or higher")
data2013$`High school or less` <- data2013$`High school graduate` + data2013$`Less than high school graduate`
data2013[,3] <- NULL
data2013[,3] <- NULL

data2013 <- data2013 %>%
  gather("category", "total", 3:5)


data2014 <- getCensus(name="acs5", 
                      vintage=2014,
                      key=census_key, 
                      vars=c("NAME", "B23006_002E",
                             "B23006_009E", "B23006_016E", "B23006_023E"), 
                      region="state:*")

data2014$state <- 2014
colnames(data2014) <- c("state", "year", "Less than high school graduate", "High school graduate", "Some college or associate's degree", "Bachelor's degree or higher")
data2014$`High school or less` <- data2014$`High school graduate` + data2014$`Less than high school graduate`
data2014[,3] <- NULL
data2014[,3] <- NULL

data2014 <- data2014 %>%
  gather("category", "total", 3:5)


data_0914 <- rbind(data2009, data2010, data2011, data2012, data2013, data2014)

# If you want to filter by another state, do so below

ct_data <- data_0914 %>%
  filter(state=="Connecticut")

ct_data2 <- ct_data %>%
  spread(category, total)

bach_base <- ct_data2[1,3]
hs_base <- ct_data2[1,4]
some_base <- ct_data2[1,5]

ct_data3 <- ct_data2
ct_data3[,3] <- ct_data2[,3] - bach_base
ct_data3[,4] <- ct_data2[,4] - hs_base
ct_data3[,5] <- ct_data2[,5] - some_base
ct_data3 <- ct_data3 %>%
  gather("category", "total", 3:5)

ct_colors <- c('#66c2a5', '#8da0cb', '#fc8d62'  )
names(ct_colors) <- unique(ct_data3$category)

gg <- ggplot(ct_data3)
gg <- gg + geom_hline(yintercept = 0)
gg <- gg + geom_line(aes(x=year, y=total, group=category, color=category))
#gg <- gg + geom_rect(aes(xmin=as.numeric(ymd("2007-12-01")), xmax=as.numeric(ymd("2010-01-01")), ymin=-Inf, ymax=Inf))
gg <- gg + annotate("rect", xmin = 2009, xmax =2010, ymin = -Inf, ymax = Inf,
                    alpha = .1)
gg <- gg + scale_color_manual(name="", values=ct_colors)
#gg <- gg + scale_y_continuous(label=percent)
gg <- gg + labs(x=NULL, y="Jobs", title="Job change in recession and recovery in Connecticut",
                subtitle="Connecticut mirrors the rest of the country for education levels.",
                caption="SOURCE: U.S. Census ACS 5-year")
gg <- gg + theme_bw(base_family="Helvetica")
gg <- gg + theme(axis.ticks.y=element_blank())
gg <- gg + theme(panel.border=element_blank())
gg <- gg + theme(legend.key=element_blank())
gg <- gg + theme(plot.title=element_text(face="bold", family="Lato Black", size=22))
gg <- gg + theme(plot.caption=element_text(face="bold", family="Lato", size=9, color="gray", margin=margin(t=10, r=80)))
gg <- gg + geom_label(data=ct_data3, aes(x=2009, y=40000, label="Recession", hjust=-.1),
                      family="Helvetica", lineheight=0.95,
                      size=5, label.size=0, color="#2b2b2b")
gg <- gg + geom_label(data=ct_data3, aes(x=2010, y=40000, label="Recovery", hjust=-.2),
                      family="Helvetica", lineheight=0.95,
                      size=5, label.size=0, color="#2b2b2b")
gg <- gg + theme(legend.position="none")
gg <- gg + theme(plot.margin = unit(c(1, 15, 1, 1), "lines"))

gg <- gg + annotation_custom(grob=textGrob("High school or less (-14k)", hjust=.5,
                                           gp=gpar(fontsize=10, 
                                                   col='#8da0cb')),
                             xmin=2015,xmax=2015,
                             ymin=-14053, ymax=-14053)
gg <- gg + annotation_custom(grob=textGrob("Bachelor's degree or higher (48k)", hjust=.50,  
                                           gp=gpar(fontsize=10, 
                                                   col='#66c2a5')),
                             xmin=2015,xmax=2015,
                             ymin=48432, ymax=48432)
gg <- gg + annotation_custom(grob=textGrob("Associate's degree or some college (21k)", hjust=.30,  
                                           gp=gpar(fontsize=10, 
                                                   col='#fc8d62')),
                             xmin=2015,xmax=2015,
                             ymin=20722, ymax=20722)
gb <- ggplot_build(gg)

gt <- ggplot_gtable(gb)

gt$layout$clip[gt$layout$name=="panel"] <- "off"
grid.draw(gt)

```

```{r us_total, fig.width=9, fig.height=7, warning=F, message=F}
# Bringing in US data. Curated from http://www.census.gov/cps/data/cpstablecreator.html

us <- read.csv("data/education_jobs_us.csv")

# Cleaning up the columns
us <- us[c("Year","Type","Totals", "Children.under.15","No.high.school.diploma",
           "High.school.or.equivalent","Some.college..less.than.4.yr.degree","Bachelor.s.degree.or.higher" )]

# Formatting the numbers to eliminate the commas, convert to numeric
us$Totals <- gsub(",", "", us$Totals)
us$Totals <- as.numeric(us$Totals)
us$Children.under.15 <- gsub(",", "", us$Children.under.15)
us$Children.under.15 <- as.numeric(us$Children.under.15)
us$No.high.school.diploma <- gsub(",", "", us$No.high.school.diploma)
us$No.high.school.diploma <- as.numeric(us$No.high.school.diploma)
us$High.school.or.equivalent <- gsub(",", "", us$High.school.or.equivalent)
us$High.school.or.equivalent <- as.numeric(us$High.school.or.equivalent)
us$Some.college..less.than.4.yr.degree <- gsub(",", "", us$Some.college..less.than.4.yr.degree)
us$Some.college..less.than.4.yr.degree <- as.numeric(us$Some.college..less.than.4.yr.degree)
us$Bachelor.s.degree.or.higher <- gsub(",", "", us$Bachelor.s.degree.or.higher)
us$Bachelor.s.degree.or.higher <- as.numeric(us$Bachelor.s.degree.or.higher)

# Years came in an awkward format, so fixing
us$year <- round(us$Year,0)

# Cleaning up and narrowing the data
us_all <- us %>%
  filter(Type=="Employed") %>%
  mutate(hs_or_less=No.high.school.diploma+High.school.or.equivalent, some_college=Some.college..less.than.4.yr.degree) %>%
  select(year, hs_or_less, some_college, Bachelor.s.degree.or.higher)

# Restructuring the data in a way that will make it graphable with ggplot2
us_all2 <- us_all %>%
  gather("type", "employed", 2:4)

us_all2$type <- gsub("Bachelor.s.degree.or.higher", "Bachelor's degree or higher", us_all2$type)
us_all2$type <- gsub("hs_or_less", "High school or less", us_all2$type)
us_all2$type <- gsub("some_college", "Associate's degree or some college", us_all2$type)

# #8da0cb blue
# #66c2a5 green
# #fc8d62 red

us_colors <- c('#66c2a5', '#fc8d62', '#8da0cb'  )

names(us_colors) <- unique(us_change$type)

#edu_colors <- c('#66c2a5', '#fc8d62', '#8da0cb')
names(us_colors) <- colnames(us_change)[2:4]

last_vals <- sapply(colnames(us_all)[2:4], function(x) last(na.exclude(us_all[,x])))
last_date <- tail(us_all$year,1) # doing this ^ wld have made it a double

us_hs_2015 <- tail(us_all$hs_or_less,1)
us_sc_2015 <- tail(us_all$some_college ,1)
us_bd_2015 <- tail(us_all$Bachelor.s.degree.or.higher,1)

gg <- ggplot(us_all2)
gg <- gg + geom_line(aes(x=year, y=employed, group=type, color=type))
gg <- gg + annotate("rect", xmin = 2008, xmax =2010, ymin = -Inf, ymax = Inf,
                    alpha = .1)
gg <- gg + scale_color_manual(name="", values=us_colors)
gg <- gg + labs(x=NULL, y="Jobs (millions)", title="Job change in recession and recovery in the U.S.",
                subtitle="For the first time, workers with a Bachelor's degree or higher make up a larger share of the workforce (36%) \nthan those with a high school diplomar or less (34%)",
                caption="SOURCE: U.S. Census CPS, Georgetown Center on Education and the Workforce")
gg <- gg + theme_bw(base_family="Helvetica")
gg <- gg + theme(axis.ticks.y=element_blank())
gg <- gg + theme(panel.border=element_blank())
gg <- gg + theme(legend.key=element_blank())
gg <- gg + theme(plot.title=element_text(face="bold", family="Lato Black", size=22))
gg <- gg + theme(plot.caption=element_text(face="bold", family="Lato", size=9, color="gray", margin=margin(t=10, r=80)))
gg <- gg + geom_label(data=us, aes(x=2008, y=max(us_all2$employed) - max(us_all2$employed)*.15, label="Recession", hjust=-.1),
                      family="Helvetica", lineheight=0.95,
                      size=5, label.size=0, color="#2b2b2b")
gg <- gg + geom_label(data=us, aes(x=2010, y=max(us_all2$employed) - max(us_all2$employed)*.15, label="Recovery", hjust=-.2),
                      family="Helvetica", lineheight=0.95,
                      size=5, label.size=0, color="#2b2b2b")
gg <- gg + theme(legend.position="none")
gg <- gg + theme(plot.margin = unit(c(1, 15, 1, 1), "lines"))
gg <- gg + annotation_custom(grob=textGrob(paste0("High school or less (", us_hs_2015, ")"), hjust=-.3,
                                           gp=gpar(fontsize=10, 
                                                   col='#8da0cb')),
                             xmin=2015,xmax=2015,
                             ymin=us_hs_2015-500, ymax=us_hs_2015-500)
gg <- gg + annotation_custom(grob=textGrob(paste0("Associate's degree or some college (", us_sc_2015, ")"), hjust=-.17,
                             gp=gpar(fontsize=10, 
                                     col='#fc8d62')),
xmin=2015,xmax=2015,
ymin=us_sc_2015, ymax=us_sc_2015)
gg <- gg + annotation_custom(grob=textGrob(paste0("Bachelor's degree or higher (", us_bd_2015, ")"), hjust=-.20,  
                             gp=gpar(fontsize=10, 
                                     col='#66c2a5')),
xmin=2015,xmax=2015,
ymin=us_bd_2015+500, ymax=us_bd_2015+500)
gg
gb <- ggplot_build(gg)

gt <- ggplot_gtable(gb)

gt$layout$clip[gt$layout$name=="panel"] <- "off"
grid.draw(gt)

```